Collaborative Visual Analysis on the Web with RCloud

### 2015-03-31 discussion.tex
fix discussion headings

{\bf Shared artifacts and artifact histories} are a central feature of
RCloud, through its shared workbooks of experiments and analyses. We
observed that hackers readily share work through RCloud. They use it to
demonstrate techniques, share results with peers and managers, and to
transfer algorithms to other groups. Artifact histories can be accessed through the notebook
tree or through GitHub's web interface.

{\bf Modularity and granularity.} The ability to partition work into
independent units (modularity) is a key to working productively in teams.
It is good if the units can be kept small, so team members can realize
benefits at least proportional to the work on the units (granularity).
RCloud's notebook and versioning capabilities allow work to be divided
into units as fine as the underlying language allows, and versioning
encourages making incremental changes at low cost and without disrupting
the work of others.

{\bf View sharing, bookmarking.} Most resources in RCloud are named
and accessed as URLs. This proved to be an effective mechanism for
sharing and for integrating analyses with external processes and systems.
It is particularly advantageous for work to be shared as URLs that
provide access to code and experiments, instead of by pasting static
screenshots into documents and presentations.

{\bf Discussion.} Annotation and commenting was another central goal.
Commenting is supported through GitHub, but our hackers found it
awkward and did not exercise it as much as we expected.
An interesting question is, to what extent should application users
be able to make annotations in published notebooks without coding
and being exposed to the hackers's view? The design of
more elaborate, integrated annotation remains as future work.

{\bf Content export} is not a capability we aimed at supporting,
and R already has many packages for this. Recently, due to popular
demand, we added user interface support for exporting plot images.
We assume that RCloud notebooks should play well in the ecosystem
of other tools, but sometimes it is difficult to know whether adding
a compatibility feature will offload complexity or bring more in.

{\bf Social-psychological incentives} and {\bf voting and ranking} are
supported through starring and forking. These mechanisms are employed often
on the platform. An obvious next step would be to enhance recommendations
using relationships discovered by static and dynamic code analysis. This
may be considered both within and across collections of scripts (the latter
being similar to VisTrails' enhanced recommendations by clustering multiple
workflows).  It seems valuable to know which packages are frequently used
together, or appear in proximity to a certain record types or data feeds.
In general, trivial or passive mechanisms to collect data for
recommendations are essential.

{\bf Group management, size and diversity.} This area needs better
support in RCloud, but is clearly important to working in teams.
We rely on external administrative processes and social conventions
to manage accounts and groups. RCloud could benefit greatly from
integrating social media to track identities and groups and to maintain
communication channels, instead of having its own isolated solution.

{\bf Curation.} Even without formal group management, users may
curate groups of related notebooks using notebook tree folders.
We found this particularly effective in collecting and distributing
training materials.

@@

{\bf Versioning.}
Some important aspects of the environment are more difficult to manage
in RCloud than in conventional systems. RCloud does not explicitly
separate the development and deployment environments. More than that,
every version of every prototype is shared by default. Although this encourages
collaboration on work in progress it also quickly exposes errors such as
misconfiguration, mismatches between versions of packages, and programming
errors that can unintentionally affect production websites. Such errors
are difficult to completely avoid, but the power of convenient sharing is often
worth it (as proven true with distributed systems in general).
Still, effective control over versioning when sharing problem needs more
attention if RCloud would be scaled up to a wider number of users lacking
informal channels to coordinate work.

@@

{\bf Caching.}
An important consideration is how and where to introduce caching
in RCloud's distributed computation model. Caching can dramatically
improve performance in a way that is otherwise transparent to
application program semantics \cite{Callahan:2006:VVM, Guo:2010:TPI}.
Though caching is usually implicit, in some environments, such as
VisTrails, programmers may also have some explicit control over cache
management. This may be desirable to ensure the repeatability of
computations that rely on volatile or unreliable data sources.
For example, the stock ticker use case is one for which the data
might temporarily become unavailable, so caching could improve
reliability and consistency of results. Alternatively, in situations
involving relatively expensive computation (for example, analysis of large
text corpora, clustering multivariate time series or deep learning algorithms)
caching may be essential to adequate performance. Currently we would
program this in RCloud by explicitly saving an analysis in a persistent database,
but it seems better to implement this capability in a general purpose associative
cache, instead of application-specific libraries. We envision cache management
being implemented in a new middle layer to be added in the future.

{\bf Security.}
It is essential to provide security in an environment for
collaboration and data publishing. To provide some
protection for RCloud workbooks in an organization's intranet,
RCloud uses an object capability model
\cite{Miller:2006:RCT}
recently added to the Rserve protocol \cite{Urbanek:2003:AFW}
that prevents unauthenticated clients from making
unauthorized calls to the RCloud runtime environment.

@@

{\bf Exploration is hard.}
R is a popular language for exploratory data analysis, and we built RCloud
as a way for data scientists to seamlessly transition from EDA to sharing
and presentation. But the combined notebook and command line interface we
designed interferes with the use of R for exploration.

@@

{\bf Discovery is hard.}
While we provided tools for searching and curating notebooks, the
search function does not help for finding methodology (and does not protect
searchers from erroneous notebooks), and the notebook tree starts to get
unwieldy with hundreds of users and many hundreds of notebooks. In addition,
our choice to retain notebooks even when they have been deleted 
exacerbates the situation, since obsolete and broken notebooks stay around
forever.

@@

{\bf Collaboration is hard.}
While forking provides a simple way to continue someone else's work, working
in teams is a lot more complicated. The forking feature is popular but it
poses a problem, too, as dozens of versions of a notebook proliferate.

>>>>

\paragraph*{Shared artifacts and artifact histories} are a central feature of
RCloud, through its shared workbooks of experiments and analyses. We
observed that hackers readily share work through RCloud. They use it to
demonstrate techniques, share results with peers and managers, and to
transfer algorithms to other groups. Artifact histories can be accessed through the notebook
tree or through GitHub's web interface.

\paragraph*{Modularity and granularity} The ability to partition work into
independent units (modularity) is a key to working productively in teams.
It is good if the units can be kept small, so team members can realize
benefits at least proportional to the work on the units (granularity).
RCloud's notebook and versioning capabilities allow work to be divided
into units as fine as the underlying language allows, and versioning
encourages making incremental changes at low cost and without disrupting
the work of others.

\paragraph*{View sharing, bookmarking} Most resources in RCloud are named
and accessed as URLs. This proved to be an effective mechanism for
sharing and for integrating analyses with external processes and systems.
It is particularly advantageous for work to be shared as URLs that
provide access to code and experiments, instead of by pasting static
screenshots into documents and presentations.

\paragraph*{Discussion} Annotation and commenting was another central goal.
Commenting is supported through GitHub, but our hackers found it
awkward and did not exercise it as much as we expected.
An interesting question is, to what extent should application users
be able to make annotations in published notebooks without coding
and being exposed to the hackers's view? The design of
more elaborate, integrated annotation remains as future work.

\paragraph*{Content export} is not a capability we aimed at supporting,
and R already has many packages for this. Recently, due to popular
demand, we added user interface support for exporting plot images.
We assume that RCloud notebooks should play well in the ecosystem
of other tools, but sometimes it is difficult to know whether adding
a compatibility feature will offload complexity or bring more in.

\paragraph*{Social-psychological incentives} and {\bf voting and ranking} are
supported through starring and forking. These mechanisms are employed often
on the platform. An obvious next step would be to enhance recommendations
using relationships discovered by static and dynamic code analysis. This
may be considered both within and across collections of scripts (the latter
being similar to VisTrails' enhanced recommendations by clustering multiple
workflows).  It seems valuable to know which packages are frequently used
together, or appear in proximity to a certain record types or data feeds.
In general, trivial or passive mechanisms to collect data for
recommendations are essential.

\paragraph*{Group management, size and diversity} This area needs better
support in RCloud, but is clearly important to working in teams.
We rely on external administrative processes and social conventions
to manage accounts and groups. RCloud could benefit greatly from
integrating social media to track identities and groups and to maintain
communication channels, instead of having its own isolated solution.

\paragraph*{Curation} Even without formal group management, users may
curate groups of related notebooks using notebook tree folders.
We found this particularly effective in collecting and distributing
training materials.

@@

\paragraph*{Versioning}
Some important aspects of the environment are more difficult to manage
in RCloud than in conventional systems. RCloud does not explicitly
separate the development and deployment environments. More than that,
every version of every prototype is shared by default. Although this encourages
collaboration on work in progress it also quickly exposes errors such as
misconfiguration, mismatches between versions of packages, and programming
errors that can unintentionally affect production websites. Such errors
are difficult to completely avoid, but the power of convenient sharing is often
worth it (as proven true with distributed systems in general).
Still, effective control over versioning when sharing problem needs more
attention if RCloud would be scaled up to a wider number of users lacking
informal channels to coordinate work.

@@

\paragraph*{Caching}
An important consideration is how and where to introduce caching
in RCloud's distributed computation model. Caching can dramatically
improve performance in a way that is otherwise transparent to
application program semantics \cite{Callahan:2006:VVM, Guo:2010:TPI}.
Though caching is usually implicit, in some environments, such as
VisTrails, programmers may also have some explicit control over cache
management. This may be desirable to ensure the repeatability of
computations that rely on volatile or unreliable data sources.
For example, the stock ticker use case is one for which the data
might temporarily become unavailable, so caching could improve
reliability and consistency of results. Alternatively, in situations
involving relatively expensive computation (for example, analysis of large
text corpora, clustering multivariate time series or deep learning algorithms)
caching may be essential to adequate performance. Currently we would
program this in RCloud by explicitly saving an analysis in a persistent database,
but it seems better to implement this capability in a general purpose associative
cache, instead of application-specific libraries. We envision cache management
being implemented in a new middle layer to be added in the future.

\paragraph*{Security}
It is essential to provide security in an environment for
collaboration and data publishing. To provide some
protection for RCloud workbooks in an organization's intranet,
RCloud uses an object capability model
\cite{Miller:2006:RCT}
recently added to the Rserve protocol \cite{Urbanek:2003:AFW}
that prevents unauthenticated clients from making
unauthorized calls to the RCloud runtime environment.

@@

\paragraph*{Exploration is hard}
R is a popular language for exploratory data analysis, and we built RCloud
as a way for data scientists to seamlessly transition from EDA to sharing
and presentation. But the combined notebook and command line interface we
designed interferes with the use of R for exploration.

@@

\paragraph*{Discovery is hard}
While we provided tools for searching and curating notebooks, the
search function does not help for finding methodology (and does not protect
searchers from erroneous notebooks), and the notebook tree starts to get
unwieldy with hundreds of users and many hundreds of notebooks. In addition,
our choice to retain notebooks even when they have been deleted 
exacerbates the situation, since obsolete and broken notebooks stay around
forever.

@@

\paragraph*{Collaboration is hard}
While forking provides a simple way to continue someone else's work, working
in teams is a lot more complicated. The forking feature is popular but it
poses a problem, too, as dozens of versions of a notebook proliferate.
